Plausibility isn't all you need: Conformal Counterfactual Explanations

### 2023-04-20 notebooks/intro.qmd
uh

M = CCE.ConformalModel(conf_model, mach.fitresult)

@@

plts = []
for (i,target) ∈ enumerate(counterfactual_data.y_levels)
    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
    Xgen = rand(sampler, nsamples)
    plt = Plots.plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
    Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=i,shape=:star,label="X|y=$target")
    push!(plts, plt)
end
Plots.plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
```

@@

niter = 100
nsamples = 100
M = CCE.ConformalModel(conf_model, mach.fitresult; likelihood=:classification_multi)

plts = []
for target ∈ counterfactual_data.y_levels
    sampler = CCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
    Xgen = rand(sampler, nsamples)
    plt = Plots.plot(M, counterfactual_data; target=target, zoom=-0.5,cbar=false)
    Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
    push!(plts, plt)
end
Plots.plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
```

@@

# Untrained Models:
models = Dict(
    :cov75 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.75)),
    :cov80 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.80)),
    :cov90 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.90)),
    :cov99 => CCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.99)),
)
```

@@

```{julia}
#| output: false
using CounterfactualExplanations.Evaluation: benchmark
bmks = []
measures = [
    CounterfactualExplanations.distance,
    CCE.distance_from_energy,
    CCE.distance_from_targets
]
for (dataname, dataset) in datasets
    bmk = benchmark(
        dataset; 
        models=deepcopy(models), 
        generators=generators, 
        measure=measures,
        suppress_training=false, dataname=dataname,
        n_individuals=10
    )
    push!(bmks, bmk)
end
bmk = reduce(vcat, bmks)
```

>>>>

M = ECCCE.ConformalModel(conf_model, mach.fitresult)

@@

plts = []
for (i,target) ∈ enumerate(counterfactual_data.y_levels)
    sampler = ECCCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
    Xgen = rand(sampler, nsamples)
    plt = Plots.plot(M, counterfactual_data; target=target, zoom=-3,cbar=false)
    Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=i,shape=:star,label="X|y=$target")
    push!(plts, plt)
end
Plots.plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
```

@@

niter = 100
nsamples = 100
M = ECCCE.ConformalModel(conf_model, mach.fitresult; likelihood=:classification_multi)

plts = []
for target ∈ counterfactual_data.y_levels
    sampler = ECCCE.EnergySampler(M, counterfactual_data, target; niter=niter, nsamples=100)
    Xgen = rand(sampler, nsamples)
    plt = Plots.plot(M, counterfactual_data; target=target, zoom=-0.5,cbar=false)
    Plots.scatter!(Xgen[1,:],Xgen[2,:],alpha=0.5,color=target,shape=:star,label="X|y=$target")
    push!(plts, plt)
end
Plots.plot(plts..., layout=(1,length(plts)), size=(img_height*length(plts),img_height))
```

@@

# Untrained Models:
models = Dict(
    :cov75 => ECCCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.75)),
    :cov80 => ECCCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.80)),
    :cov90 => ECCCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.90)),
    :cov99 => ECCCE.ConformalModel(conformal_model(clf; method=:simple_inductive, coverage=0.99)),
)
```

@@

```{julia}
#| output: false
using CounterfactualExplanations.Evaluation: benchmark
bmks = []
measures = [
    CounterfactualExplanations.distance,
    ECCCE.distance_from_energy,
    ECCCE.distance_from_targets
]
for (dataname, dataset) in datasets
    bmk = benchmark(
        dataset; 
        models=deepcopy(models), 
        generators=generators, 
        measure=measures,
        suppress_training=false, dataname=dataname,
        n_individuals=10
    )
    push!(bmks, bmk)
end
bmk = reduce(vcat, bmks)
```

### 2023-04-20 notebooks/mnist.qmd
uh

```{julia}
# Data:
n_obs = 10000
counterfactual_data = load_mnist(n_obs)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 200
activation = Flux.relu
builder = MLJFlux.@builder Flux.Chain(
    Dense(n_in, n_hidden),
    BatchNorm(n_hidden, activation),
    Dense(n_hidden, n_hidden),
    BatchNorm(n_hidden, activation),
    Dense(n_hidden, n_out),
)
# builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.2, σ=activation)
# builder = MLJFlux.MLP(
#     hidden=(
#         n_hidden,
#         n_hidden,
#         n_hidden,
#     ), 
#     σ=activation
# )
α = [1.0,1.0,1e-2]

@@

# Joint Energy Model:
𝒟x = Uniform(0,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10
)
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    batch_size=batch_size,
    finaliser=x -> x,
    loss=Flux.Losses.logitcrossentropy,
    jem_training_params=(
        α=α,verbosity=10,
        # use_gen_loss=false,
        # use_reg_loss=false,
    ),
    sampling_steps=20,
    epochs=epochs,
)

@@

```{julia}
cov = .90
conf_model = conformal_model(jem; method=:adaptive_inductive, coverage=cov)
mach = machine(conf_model, X, labels)
fit!(mach)
M = CCE.ConformalModel(mach.model, mach.fitresult)
```

@@

# CCE:
λ=[0.0,1.0]
temp=0.01

# Generate counterfactual using CCE generator:
generator = CCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(),
)
ce_conformal = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Generate counterfactual using CCE generator:
generator = CCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
)
ce_conformal_jsma = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

@@

ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
_names = ["Wachter", "CCE", "JSMA", "CCE-JSMA"]
for x in zip(ces, _names)
    ce, _name = (x[1],x[2])
    x = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "cce_mnist.png"))
```

@@

# Measures:
measures = [
    CounterfactualExplanations.distance,
    CCE.distance_from_energy,
    CCE.distance_from_targets,
    CounterfactualExplanations.validity,
]
```

>>>>

```{julia}
function pre_process(x; noise::Float32=0.03f0)
    ϵ = Float32.(randn(size(x)) * noise)
    x = @.(2 * x - 1) .+ ϵ
    return x
end
```

```{julia}
# Data:
n_obs = 10000
counterfactual_data = load_mnist(n_obs)
X, y = CounterfactualExplanations.DataPreprocessing.unpack_data(counterfactual_data)
X = pre_process.(X)
X = table(permutedims(X))
labels = counterfactual_data.output_encoder.labels
input_dim, n_obs = size(counterfactual_data.X)
n_digits = Int(sqrt(input_dim))
output_dim = length(unique(labels))
```

@@

```{julia}
# Model parameters:
epochs = 100
batch_size = minimum([Int(round(n_obs/10)), 128])
n_hidden = 32
activation = Flux.relu
# builder = MLJFlux.@builder Flux.Chain(
#     Dense(n_in, n_hidden, activation),
#     Dense(n_hidden, n_hidden, activation),
#     Dense(n_hidden, n_hidden, activation),
#     # BatchNorm(n_hidden, activation),
#     # Dense(n_hidden, n_hidden),
#     # BatchNorm(n_hidden, activation),
#     Dense(n_hidden, n_out),
# )
builder = MLJFlux.Short(n_hidden=n_hidden, dropout=0.2, σ=activation)
# builder = MLJFlux.MLP(
#     hidden=(
#         n_hidden,
#         n_hidden,
#         n_hidden,
#     ), 
#     σ=activation
# )
α = [1.0,1.0,1e-2]

@@

# Joint Energy Model:
𝒟x = Uniform(-1,1)
𝒟y = Categorical(ones(output_dim) ./ output_dim)
sampler = ConditionalSampler(
    𝒟x, 𝒟y, 
    input_size=(input_dim,), 
    batch_size=10
)
jem = JointEnergyClassifier(
    sampler;
    builder=builder,
    batch_size=batch_size,
    finaliser=x -> x,
    loss=Flux.Losses.logitcrossentropy,
    jem_training_params=(
        α=α,verbosity=10,
        # use_gen_loss=false,
        # use_reg_loss=false,
    ),
    sampling_steps=20,
    epochs=epochs,
)

@@

```{julia}
cov = .95
conf_model = conformal_model(jem; method=:adaptive_inductive, coverage=cov)
mach = machine(conf_model, X, labels)
fit!(mach)
M = ECCCE.ConformalModel(mach.model, mach.fitresult)
```

@@

# ECCCE:
λ=[0.0,1.0]
temp=0.01

# Generate counterfactual using ECCCE generator:
generator = ECCCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=Flux.Optimise.Adam(),
)
ce_conformal = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

# Generate counterfactual using ECCCE generator:
generator = ECCCEGenerator(
    λ=λ, 
    temp=temp, 
    opt=CounterfactualExplanations.Generators.JSMADescent(η=1.0),
)
ce_conformal_jsma = generate_counterfactual(
    x, target, counterfactual_data, M, generator; 
    decision_threshold=γ, max_iter=T,
    initialization=:identity,
    converge_when=:generator_conditions,
)

@@

ces = [ce_wachter, ce_conformal, ce_jsma, ce_conformal_jsma]
_names = ["Wachter", "ECCCE", "JSMA", "ECCCE-JSMA"]
for x in zip(ces, _names)
    ce, _name = (x[1],x[2])
    x = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(www_path, "cce_mnist.png"))
```

@@

# Measures:
measures = [
    CounterfactualExplanations.distance,
    ECCCE.distance_from_energy,
    ECCCE.distance_from_targets,
    CounterfactualExplanations.validity,
]
```

### 2023-04-20 paper/paper.tex
title
uh

\title{Conformal Counterfactual Explanations}

@@

Penalizing the set size in this way is in principal enough to train efficient conformal classifiers \citep{stutz2022learning}. As we explained above, the set size is also closely linked to predictive uncertainty at the local level. This makes the smooth penalty defined in Equation~\ref{eq:setsize} useful in the context of meeting our objective of generating plausible counterfactuals. In particular, we adapt Equation~\ref{eq:general} to define the baseline objective for Conformal Counterfactual Explanations (CCE):

@@

In order to generate prediction sets $C_{\theta}(f(\mathbf{Z}^\prime);\alpha)$ for any Black Box Model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$. Arguably, data is typically abundant and in most applications practitioners tend to hold out a test data set anyway. Our proposed approach for CCE therefore removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.

>>>>

\title{Plausibility isn't all you need: Conformal Counterfactual Explanations}

@@

Penalizing the set size in this way is in principal enough to train efficient conformal classifiers \citep{stutz2022learning}. As we explained above, the set size is also closely linked to predictive uncertainty at the local level. This makes the smooth penalty defined in Equation~\ref{eq:setsize} useful in the context of meeting our objective of generating plausible counterfactuals. In particular, we adapt Equation~\ref{eq:general} to define the baseline objective for Conformal Counterfactual Explanations (ECCCE):

@@

In order to generate prediction sets $C_{\theta}(f(\mathbf{Z}^\prime);\alpha)$ for any Black Box Model we merely need to perform a single calibration pass through a holdout set $\mathcal{D}_{\text{cal}}$. Arguably, data is typically abundant and in most applications practitioners tend to hold out a test data set anyway. Our proposed approach for ECCCE therefore removes the restriction on the family of predictive models, at the small cost of reserving a subset of the available data for calibration.
